Sex-Specific Evolution of the Meiotic Recombination Rate

### 2020-06-17 doc/MS_v3.Rmd
continued work on intro

Meiosis can be reduced to the expression of (2n -> 4n -> 2n -> 1n) which tracks the duplication of a diploid genome into haploid cell products. The meiotic program relies on crossovers and the process of recombination to ensure the correct separation of chromosomes. The total number of COs per cell (at the 4n stage) is equal to the genome wide recombination rate (gwRR). **The genome wide recombination rate is a cell based metric for genome wide recombination rate is tightly connected to organisms fitness through correct chromosome segregation and fertility.**

**at the fine scale / in regards to genetic variation**, In the indriect manner, the recombination rate regulates populations’ response to selection, and determine the fate of novel mutations by transferring beneficial mutations onto novel genetic backgrounds or by breaking linkage of negative mutations from beneficial genetic backgrounds (cite).  This process shapes the genomic patterns of genetic variation, with high recombination areas of genome having more nucleotide variation while areas of low recombination have decreased genetic variation (Begun Aquadro, Nachman, Payseur).

**but back to the genome-wide recombination rate**, The process of chromosome segregation is integral to the process of producing haploid cells and can be considered a constraint in regards to the evolution. When the genome wide recombination rate is framed within the single cell context it is closely connected to the  contraints which are imposed at the single cell level. Reflecting the importance of it's role in chromosome segreation is that the  number of haploid pairs is strong predictor for genome wide RR across large species(taxa) / divergence scales [@ottoPaysuer2019; @stapley_variation_2017]. (the number of crossovers per chromosome are highly conserved)<However, whie the rate of recombination is highly variable, the count of crossovers per chromosome is conserved between 1 to 3 (cite), suggesting that such thresholds/constraints may be behind the overall patterns.>

**connection of single cell based gwRR to thresholds imposed at the single cell level** (why is gwRR the better scale for evolutionary patterns?)

@@

@@

@@

>>>>

Meiosis can be reduced to the expression (2n -> 4n -> 2n -> 1n) which reflects the process diploid genome producing haploid cell products. In most species the meiotic program relies on crossovers to ensure the correct separation of chromosomes at the reductional division (4n to 2n) [Cahoon]. The total number of COs per cell (at the 4n stage) is equal to the genome wide recombination rate (gwRR). Additioanlly, the gwRR can regulate a populations’ response to selection, and determine the fate of novel mutations by transferring beneficial mutations onto novel genetic backgrounds or by breaking linkage of negative mutations from beneficial genetic backgrounds (cite).  This process shapes the genomic patterns of genetic variation, with high recombination areas of genome having more nucleotide variation while areas of low recombination have decreased genetic variation [@begun1992; @nachman2012]. **Because of it's role in facilitating chromosomes disjunction, the genome wide recombination rate can be thought of as a cell-based metric which is tightly connected to organisms fitness and fertility.**

@@

The process of chromosome segregation is integral to the process of producing haploid gametes and can be considered a constraint in regards to the evolution of the genome-wide recombination rate.  Ensuring at least one crossover per chrm while minimizing the rates of DNA damage and ectopic recombination (cite) are considered to be the lower and upper thresholds for a genome-wide recombiantion rate [@nagaoka2012]. The number of haploid pairs of chromosomes is strong predictor for genome wide recombination rate across large divergence scales [@ottoPaysuer2019; @stapley_variation_2017]. Yet, within these thresholds the genome wide recombination rate still varies across species and individuals. Sex is one of the most notable axes of this form of variation. 
(examples, Humans 1.7 female to male
alligators female to male ratio is X, temperature dependent sex determination)

<2.	Gap is due to Lack of comparable measures>
An understanding of how sex shapes the evolution of recombination cannot be achieved with available data. Comprehensive comparisons of female and male recombination rates usually come from outbred populations humans [@Kong2004;, @Kong2008; @Kong2014; @halldorsson2019], dog,  cattle [@ma2015_cattle; @Shen2018_cattle], (other sheep?) sheep [@johnston2016_soay], and mouse (CC cite) in which the role of sex is confounded with the contributions of genetic variation. 
(other gaps in data )

<3.	Open Q, no first principle>
Under anisogamy where gametogenesis is modified to produce distinct gametes (cite Gorelick?), the meiotic program is the same and there are no first principles which would predict the evolution of sexually dimorphic recombination rates. Yet heterochiasmy is commonly observed in dioecious species, suggesting that other meiotic traits which distinguish the gametes and their meiotic programs, for example symmetrical vs asymmetrical cell division, may impose selection for sexually dimorphic recombination rates. Although it is clear that the relationship between female and male recombination rates can differ among species, comparisons between and within closely related species are missing. Direct contrasts between females and males across a common, diverse set of genomic backgrounds would reveal whether the recombination rate evolves differently in the sexes.

<4.	Additional metrics (more nuanced)>
Initial focus/evidence on heterochiasmy was primarily focused on which sex had more recombination – the field has advanced since it’s first documentation (cite morgan 1914). Recently two specific features have been highlighted as conserved patterns of heterochiasmy: @CahoonLibuda2019 show that in species with meiotic chromatin (SC axis length) quantified for both sexes, longer chromosome axis generally has higher recombination rates. 
In @sardell_sex_2020, a survey of 51 species shows there are conserved broad scale recombination landscape differences. Generally males have telomere-bias crossover placement and females have more uniform placement.

These observations raise the question of how conservation in sexually dimorphic patterns are maintained with the gwRR and it’s decomposed traits.

-----------

@@

The House mouse species complex arose from a recent radiation providing an opportunity to interrogate natural variation at short evolutionary scales, potentially narrowing the mutational space for identifying variants which would cause gwRR variation. Add divergence ranges. divergence spanning 0.5 mya to 5 mya  (cite / Geraldes et al 2008 and Geraldes et al 2010).

Classical lab strains of mice have generated extensive knowledge central to meiosis and outcomes on recombination through extensive studies on the genes involved in the meiotic recombination pathway (cite) and previous crosses for understanding the genetic architecture of recombination rate variation @dumont2011house; @murdoch2010; @Wang2017island].
Additionally due to their global distribution allows us to compare samples of natural diversity from a broad geographic range.  (wild derived inbred strains – will capture genetic diversity not represented in classical lab strains.)

**Wild derived inbred strains generate the best comparison of females and males, besides the sex chromosomes, the mouse for each genome are almost identical**. Inbred strains enable one of the most direct comparison of both female and male gametogenesis and enable us to conclude that sex is a primary factor.

House mouse is well suited for single cell cytology approaches (due to large cell sizes). The single cell level allows integration of data at a closer connection to the molecular pathway and meiotic program. In addition to the quantification of gwRR. This study quantifies precursors to crossovers, double strand breaks (DSBs), and meiotic chromosome morphology across stages of meiosis which would not be accessible to with genetic linkage data and crosses.

@@

Main messages: these results are centered/focused on a unique evo comparison 
Advantages / importance of using multiple strains from and multiple subspecies
-importance / advantage of sampling for multiple geographic locations.
Chromosome morphology / general recombination landscape
-results for general landscape – conserved sex differences  and

---------------
OUR results for genome wide recombination rate (variation) we observed highly distinct evolutionary trajectories between males and females. This is due to rapid male specific evolution in the gwRR from two strains in musculus and 1 strain in molossinus. (this pattern of rapid evolution was not observed in females). We observed up to a 30% difference (translating into approximately ~7 more crossovers per cell), a surprising amount considering the short evolutionary time scale.

Conserved patterns of the recombination landscape of the single chromosome level – reflect constraints  At the single chromosome level – (crossover number is conserved – suggesting processes regulating the rec landscape at this scale might face similar constraints

------
